---
title: "How Does a Bike-Share Navigate Speedy Success Case study"
author: "Skillarchie davis"
date: "6/1/2021"
output: html_document
---
## Step 1: Import your data

The data is own by the city of Chicago was colleted by Motivate International Inc.

You can learn more about the data integrity here:
https://www.divvybikes.com/data-license-agreement

You can retrieve the dataset from here:
https://divvy-tripdata.s3.amazonaws.com/index.html

Run the code below to read in the file 'hotel_bookings.csv' into a data frame: 
```{r}
library(readr)
Bike_rental <- read_csv("C:/Users/skill/Desktop/Projects/Datasets/202104-divvy-tripdata.csv")
#View(X202104_divvy_tripdata)
```

```{r}
library("lubridate")
library(jsonlite)
library(ggplot2)
library(dplyr)
library(tidyr)
library(tidyverse)
# text mining library
library(tidytext)
library(tm)
# coupled words analysis
library(widyr)
# plotting packages
library(igraph)
```

## Step 2: Refresh Your Memory

Preview the dataset with head(), glimse, summary and colnames functions

```{r}
head(Bike_rental)
```

```{r}
glimpse(Bike_rental)
```

```{r}
colnames(Bike_rental)
```

```{r}
summary(Bike_rental)
```


## Step 3: Check for inconsistency in the Data. 

This may inlude null values, incorrect input and wrong data type.

### checking how many rows are incomplete
```{r}
nrow(Bike_rental[!complete.cases(Bike_rental),])
```


### checking how many NA per column
```{r}
apply(Bike_rental,2,function(x) sum(is.na(x))) 
```


The result from the apply function made me decide to investigate the incomplete columns by using histogram and boxplot.

### The use of visualisation to see extent of the outliers within each column.
```{r}
hist(Bike_rental$start_lat)
hist(Bike_rental$end_lat)
hist(Bike_rental$start_lng)
hist(Bike_rental$end_lng)
```


```{r}
boxplot(Bike_rental$start_lat)
boxplot(Bike_rental$end_lat)
boxplot(Bike_rental$start_lng)
boxplot(Bike_rental$end_lng)
```

## Step 4 : Remove NA's from the dataset

After visualising the inconsistency in the data I decided to removed the NA's 
```{r}
Bike_rental <- na.omit(Bike_rental)
```

## Step 5 : Check to see if NA's values were removed

```{r}
apply(Bike_rental,2,function(x) sum(is.na(x))) 
```


## The summary of all columns after data mining and cleaning.
```{r}
summary(Bike_rental)
```

## Step 6: Formatting the data for analysis

This include adding extra columns and also deleting coulums that are not needed.

```{r}
library(chron) # library to convert character back to times data type

# stripe the Times format from the Datetime data type
Bike_rental = Bike_rental %>% mutate(start_time = strftime(Bike_rental$started_at, tz = "GMT", format = "%H:%M:%S"),end_time = strftime(Bike_rental$ended_at, tz = "GMT", format = "%H:%M:%S"))

# Convert from character back to time 
Bike_rental = Bike_rental %>% mutate(start_time = chron(times = start_time),end_time = chron(times = end_time))


# create a new column ride_length by substrating starting time from end time
Bike_rental = Bike_rental %>% mutate(ride_length = as.numeric(end_time - start_time))


Bike_rental = Bike_rental %>% mutate(ride_length = format(as.POSIXct((ride_length) * 86400, origin = "1970-01-01", tz = "UTC"), "%H:%M:%S"))# convert to hms format but datatype is character

Bike_rental = Bike_rental %>% mutate(ride_length = chron(times = ride_length)) # convert back to time

class(Bike_rental$ride_length)
```


## Converting the date to weekdays

This will help me get better insights when I start to analyze the data

```{r}
Bike_rental$day_of_week_start <- weekdays(Bike_rental$started_at)
Bike_rental$day_of_week_end <- weekdays(Bike_rental$ended_at)

```


## Converting the date to Months
```{r}
Bike_rental$Start_Month <- months(Bike_rental$started_at)
Bike_rental$End_Month <- months(Bike_rental$ended_at)
```

## Converting the date to Years
```{r}
Bike_rental$Year <- year(Bike_rental$started_at)
View(Bike_rental)
```


## Cleaning Bad Data 
```{r}
Bike_rental <- Bike_rental[!(Bike_rental$start_station_name == "HQ QR" | Bike_rental$ride_length<0),]
```


### Reassign to the desired values 
```{r}
Bike_rental <-  Bike_rental %>% 
  mutate(member_casual = recode(member_casual,"Subscriber" = "member","Customer" = "casual"))

```

## Step 7: Analyze the data for insights


### Descriptive analysizing the ride_length column
```{r}
min(Bike_rental$ride_length)#shortest ride
```


```{r}
max(Bike_rental$ride_length)#longest ride
```

```{r}
median(Bike_rental$ride_length)#midpoint number in the ascending array of ride lengths
```


```{r}
mean(Bike_rental$ride_length)
```



# Compare members and casual users
```{r}
aggregate(Bike_rental$ride_length ~ Bike_rental$member_casual, FUN = mean)
aggregate(Bike_rental$ride_length ~ Bike_rental$member_casual, FUN = median)
aggregate(Bike_rental$ride_length ~ Bike_rental$member_casual, FUN = max)
aggregate(Bike_rental$ride_length ~ Bike_rental$member_casual, FUN = min)

```

### Checking the most common days and months
```{r}
library(modeest) # Conatins the mfv() function that will be used to calculate mode.
casual_rider <- subset(Bike_rental, member_casual == "casual")
membership_rider <- subset(Bike_rental, member_casual == "member")
```

### Most frequent days and month that for a casual rider
```{r}
mfv(casual_rider$day_of_week_start) # mode function. Returns most frequent value.
mfv(casual_rider$day_of_week_end)
mfv(casual_rider$Start_Month)
```

### Most frequent days and month that for a Membership rider
```{r}
mfv(membership_rider$day_of_week_start)
mfv(membership_rider$day_of_week_end)
mfv(membership_rider$Start_Month)
```

### Calculating the average ride_length for members and casual riders
```{r}
Bike_rental %>%
	filter( member_casual == "casual" | member_casual =="member") %>%
	group_by(member_casual) %>%
	summarise(mean_by_rider_type = mean(ride_length))
```


### Calculating the average ride_length for week days for members and casual users
```{r}
Bike_rental %>%
	mutate(weekday = wday(started_at, label = TRUE)) %>%
	group_by(member_casual,weekday) %>%
	summarise(mean_by_rider_type = mean(ride_length),.groups = 'drop') 
```


## Calculate the number of rides for users by day_of_week 
```{r}
Bike_rental %>% 
  mutate(weekday = wday(started_at, label = TRUE)) %>% 
  group_by(member_casual, weekday) %>% 
  summarise(number_of_rides = n()
            ,average_duration = mean(ride_length), .groups = 'drop') %>%
  arrange(member_casual, weekday) 
```


## Let's visualize the number of rides by type of bicycle separated by rider type
```{r}
Bike_rental %>% 
  mutate(weekday = wday(started_at, label = TRUE)) %>% 
  group_by(member_casual, rideable_type, weekday) %>% 
  summarise(number_of_rides = n()
            ,average_duration = mean(ride_length), .groups = 'drop') %>%
  arrange(member_casual, weekday) %>% 
  ggplot(aes(x =weekday, y = number_of_rides, fill = rideable_type)) +
  labs(title = "Number of  of Rides by type of bicycle separated by rider type", x = "Week Days", y = "Number of Rides")+
  geom_col(position = "dodge")+ 
  facet_wrap(~member_casual)
ggsave("Number of  of Rides by type of bicycle separated by rider type.png")
```



## Let's visualize the number of rides by rider type
```{r}
Bike_rental %>% 
  mutate(weekday = wday(started_at, label = TRUE)) %>% 
  group_by(member_casual, weekday) %>% 
  summarise(number_of_rides = n()
            ,average_duration = mean(ride_length), .groups = 'drop') %>% 
  arrange(member_casual, weekday)  %>% 
  ggplot(aes(x = weekday, y = number_of_rides, fill = member_casual)) +
  labs(title = "Number of Riders Per Day Group by Type of Customer", x = "Week Days", y = "Number of Rides")+
  geom_col(position = "dodge")
ggsave("Number of Riders Per Day Group by Type of Customer.png")
```

## Let's create a visualization for average duration
```{r}
Bike_rental %>% 
  mutate(weekday = wday(started_at, label = TRUE)) %>% 
  group_by(member_casual, weekday) %>% 
  summarise(number_of_rides = n()
            ,average_duration = mean(ride_length), .groups = 'drop') %>% 
  arrange(member_casual, weekday)  %>% 
  ggplot(aes(x = weekday, y = average_duration, fill = member_casual)) +
  labs(title = "Average Duration Per Week Day Group by Type of Customer", x = "Week Days", y = "Average Duration") +
  geom_col(position = "dodge")
ggsave("Average Duration Per Week Day Group by Type of Customer.png")
```



## Exporting the data

```{r}
write.csv(Bike_rental,file = file.choose(new = T))
```










